---
# Rmarkdown header. it use Use the YAML syntax. 
# Just copy the ---YAML block --- and change title, subtyle, author and date
title: "Task 1 Introduction" 
subtitle: "People Analytics"
author: "Jesús García Romanos"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true

    code_folding: hide

# store the output in the subdirectory reports
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_dir=file.path(dirname(inputFile),"reports" )) }) 
---

# Execute your Rnotebook

- **Run Code by line**: Use `Ctrl + Enter` (Windows) or `Cmd + Enter` (Mac) to run lines of code directly from the chunks (scripts contained in the grey areas).
- **knit the complete Rnotebook**: press the 🧶 Knit buttom at the top 

# Setup

## Define directories

We will organize your project files into specific folders within the main project directory. ***If they do not exist, create the following directories and move this Rnotebook —task1_introduction.Rmd— to the notebooks directory ***.

* **data** to store the data sources and data created. 
* **notebooks** for storing R notebooks.

This directory will be created the first time you knit (Execute) this Rnotebook.

* **reports** to store the web (HTML) output files generated when knitting.

## Load Packages

- Base R has basic functions (commands), but we can extend R with +22,000 packages. 

- During the course, we will make extensive use of **tidyverse**, a collection of R packages designed to work together for data science, using a consistent and beginner-friendly style. So we will load it into all our notebooks. 

- **here()** package helps us to work neatly in the project working directory when I move Rnotebooks between computers

### Do not worry to see

    ## ── Conflicts ────────────────────────────────────────── tidyverse_conflicts() ──
    
- It just indicates that when you load the library, you mask some functions with the same name that were already loaded. 

```{r setup}

# ==============================================================================
# library(tidyverse) loads the following packages:
# ==============================================================================

# 1. ggplot2   -> Data Visualization (The "Grammar of Graphics")
# 2. dplyr     -> Data Manipulation (filter, select, mutate, summarize)
# 3. tidyr     -> Data Tidying (reshaping long/wide formats)
# 4. readr     -> Data Import (fast reading of .csv, .tsv, and fwf)
# 5. purrr     -> Functional Programming (replacing loops with map functions)
# 6. tibble    -> Modern Data Frames (cleaner printing and stricter subsetting)
# 7. stringr   -> String Manipulation (all functions start with str_)
# 8. forcats   -> Factor Handling (tools for categorical variables)
# 9. lubridate -> Date/Time Wrangling (parsing and math with temporal data)

# ==============================================================================

library(tidyverse)
# ==============================================================================
# Masking happens when two packages contain functions with the same name.
# When you load a package, its functions take priority over previously loaded ones.
# The conflicts message simply tells you which version R will use by default.
# ==============================================================================

library(here)
knitr::opts_knit$set(root.dir = here::here())

```
**Are you feeling a bit lost here?** 


- Think of R like a TV with multiple streaming apps installed and a single remote control with a button labeled **filter**. 
  Base R comes with a built-in app called stats, which includes a function called ***filter()*** that runs when you press the **filter** button. 
  
  When you load dplyr, you install another app that also has a function ***filter()*** to run when you press the **filter** button.
  Now you have two programs named ***filter()***. When you press the **filter** button, R has to decide which apps to use.
  
  The conflict message is R saying, as the ***filter()*** from dplyr was the latest to load:
  “From now on, when you press **filter**, I’ll use the one from dplyr instead of the one from stats.” 
  
  If you really want the other one, you can say:
  
  *stats::filter()*
  
  That’s like saying:
  “No, I meant the ***filter()*** function from the built-in app.”  
  

In future tasks, the setup chunk will be **hidden** from the notebook output by adding *,include=FALSE* to the chunk name, however R will load packages and mask functions without displaying the above messages.

# where am I ?

To know in which directory I am working I use *getwd()*
```{r}
getwd()
```

# what files do I have in my project?

To know which files I have in my project directory I use *dir()*
```{r}
dir()
```


# Exploring R Commands


### Define and show values

vales are simple variables to be used in any moment. 

- Assign a name to a variable. **<-** and **=** are used indistinctly 

      myname = "Jesús"
      myname <- "Jesús"
    
- **cat()**  function to concatenate strings and display them
```{r define values}
my_name = "Jesús"
myname <- "Jesús"

city_of_birth = "Avila"

cat(my_name, "was born in ", city_of_birth)

numbers=0:9
cat("the numbers are:", numbers)
```

### Define and show data

Normally you work with structured data like data frames (tibble in tidyverse) where you define structure information, like the full name wich has to parts, a name and a surname, and you can work with the full name, or with each part separately depending on your needs

- **print()**  function to print structure data.  cat() cannot handle this type of data, so we use print() instead.
```{r Define data}

myfullname = tibble(name="Jesús",
                    surname="García Romanos")

print(myfullname)
print(myfullname["surname"])


cat(myfullname$name,myfullname$surname)


```
