---
# Rmarkdown header. it use Use the YAML syntax. 
# Just copy the ---YAML block --- and change title, subtyle, author and date
title: "Task 3 data cleaning" 
subtitle: "People Analytics"
author: "Jesús García Romanos"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide

# store the output in the subdirectory reports
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_dir=here::here("reports")) }) 
---

# Setup

```{r setup, include=FALSE}
#this chunk is hidden from the notebook output by adding ,include=FALSE to the chunk name.

library(tidyverse) 

library(here)

```

### Read df_attrition

✍️ your task

- read df_atrition.rds


```{r}



```


# Add new employee


### Read new employee data

🔦 The Research & Development department has sent us data from a new employee, that we want to append to our df_attrition database, The format of the data is whit the names of each column (variables), followed but its value. Those variables that the user didn't know wrote "**dontknow**".

✍️ your task

- Download the file employee_record.txt from Moodle to the data directory.
- Review the file and read it into df_new_employee
- Compare the structure of the new record with that of the df_attrition

📝️ R commands
    - **read_tsv()**: read tab separated files. 
      💡 It is a particular case of **read_delimin()** is a command to read plain texts, which allows to read plain text with any separator.  

```{r read_new_employee}

```
### Transform new employee data

🔦 The  file employee_record.txt has each employee variable in a row, so we need to transpose rows by columns so we can add to df_attrition data frame. we do so with pivot_wider, which transforma rows into columns. 

✍️ your task

- transpose df_new_employee and convert types to match the structure of df_attrition
- Add the new employee record to df_attrition

📝️ R commands
    - pivot_wider() transform rows into columns  
    - type_convert() change the type of columns, to the more likely one.  
      i.e. if every value is a number transforms character into numeric. 
    - bind_rows() allows to add new employees to an existing data frame  
    
```{r transform_new_employee}


```

### Merge new employee into df_attrition

✍️ your task

- Add the new employee record to df_attrition

📝️ R commands

    - bind_rows() allows to add new employees to an existing data frame 

```{r bind_new_employee}


```

# Data cleaning

## Filling missing values

### show missing values

✍️ your task

- show employees with missing values, subseting columns EmployeeNumber, Age, and any column with missingvalues

📝️ R commands

- **where(~ any(is.na(.)))**: allows to select columns with missing values.
- **pull()**: convert a variable into a value vector
- **paste()**: concatenate a vector
    

```{r show_NA}

```

### Fill missing values

✍️ your task

- Fill missing Monthly income values with the average of all employees with the same Job role, level, and department.
- Fill the missing values of TotalWorkingYears, estimating their total working years using their years at the current company plus an adjustment based on how many companies they have worked for.

📝️ R commands

- **group_by()**: group observations (employees) by several variables, so you can estimate statistics by each unique value of those groups.
- **ungroup()**: remove grouping to avoid future mishandling of groups


```{r fill_NA}

```

## Remove duplicates rows

✍️ your task

- Identify duplicated rows using  *group_by* and *filter*.
- Identify duplicated rows using  **duplicate(.)**.

📝️ R commands

- **duplicate(.)**: Identify the duplicated rows, omitting first duplicated observation. 
- **distinct()**: Removing duplicates rows.



```{r}


```

# Save data

✍️ your task

- save df_attrition into disk for future use.

```{r}


```

**Congratulations, we have finished our Data Cleaning!**
